---
title: "SQL, the Sequel : Data Wranglers DC : August 2014 Meetup"
output:
  knitrBootstrap::bootstrap_document:
    title: "SQL, the Sequel : Data Wranglers DC : August 2014 Meetup"
    theme: flatly
    highlight: github
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

SQL, the Sequel : Data Wranglers DC : August 2014 Meetup
========================================================
**Author:**  [Ryan B. Harvey](http://datascientist.guru)

**Created:** 2014-07-17 / **Last Updated:** 2014-07-18

**Contents:**
- The `sqldf` package: SQL on dataframes in R
- Loading some test data
- Querying and subsetting that test data
- Example of a complex query

**Package Dependencies:**
- **[sqldf](http://cran.r-project.org/package=sqldf):** Perform SQL Selects on R Data Frames. R package version 0.4-7.1. Author: G. Grothendieck (2014) [more info & source](https://code.google.com/p/sqldf/)

**References:**
- This RPub was created for a talk to the Data Wranglers DC meetup group on August 6, 2014.
  - Talk information: [http://www.meetup.com/Data-Wranglers-DC/events/177269432/](http://www.meetup.com/Data-Wranglers-DC/events/177269432/)
  - This is only a portion of the talk, which included database information, SQL code, and an iJulia notebook too.
  - My code, slides and the sample data are all available on Github. [https://github.com/nihonjinrxs/dwdc-august2014](https://github.com/nihonjinrxs/dwdc-august2014)

The `sqldf` package is an R package that creates tables in a local SQLite database, runs the requested SQL query on those tables, and then drops the tables from the SQLite database.  Although this can be slower than subsetting a data frame directly, in cases where the data frame is very complex and conditions get difficult to parse, SQL can provide a straightforward, declarative way of requesting a particular data subset.

If the `sqldf` package doesn't exist in your environment, install the package.  Once you have it, require it.
```{r}
# install.packages("sqldf")
require(sqldf)
```

Note that the `sqldf` package will use SQLite as a default, unless a different database driver package is currently loaded (i.e., If we would like `sqldf` to use a different database product, we can tell it which driver to use by specifying that in the options statement.
```{r}
# options(sqldf.driver = "SQLite")
```

For example, if you wanted to use some of the functions available in PostgreSQL, you could use that as the driver for `sqldf` by loading the `RPosgreSQL` package first.  More details on this are at [How does one use sqldf with PosrgreSQL?](https://code.google.com/p/sqldf/#12._How_does_one_use_sqldf_with_PostgreSQL?).

Load some data to work with
---------------------------
For this example, we'll use the [Basic Business Licenses data](http://data.dc.gov/Main_DataCatalog.aspx?id=1520) from the [DC government open data site](http://data.dc.gov/).

The data is available as a zip file, so we'll download it if it's not already available, and then unzip it if it hasn't been already.

```{r}
url = "http://data.octo.dc.gov/feeds/bbl/bbl_current_csv.zip"
data_dir = "../data/"
current_dir = getwd()
setwd(data_dir)
local_zipped = "bbl_current_csv.zip"
local_data = "bbl_current_csv.csv"
if (!(file.exists(local_data) && file.exists(local_zipped))) {
  download.file(url, destfile=local_zipped, method="curl")
  fileDownloadedDate <- date()
  paste("File downloaded at ", fileDownloadedDate)
}
if (!file.exists(local_data)) {
  unzip(local_zipped)
}
bbl <- read.csv(local_data)
setwd(current_dir)
```

Now that we have the data frame `bbl`, let's take a look at what's in it.

```{r}
head(bbl)
summary(bbl)
```

SQL Queries on Data Frames
--------------------------

Now, let's do a few queries on it just to get the hang of using `sqldf`.

```{r}
sqldf("select * from bbl where BUSINESSIMPROVEMENTDISTRICT = 'DOWNTOWN BID'")
```

```{r}
sqldf("select * from bbl where WARD = 6")
```

The `sqldf()` function returns a data frame, so we can store it in another variable for use in a plot, for example.

```{r}
ward6 <- sqldf("select LICENSECATEGORY, count(*) as LICENSECOUNT from bbl where WARD = 6 group by LICENSECATEGORY")
ward6
```

```{r fig.width=10, fig.height=6}
par(las=2, mar=c(2, 12, 2, 2)) # Labels perpendicular to axis, wide left margin
barplot(ward6$LICENSECOUNT, names.arg=ward6$LICENSECATEGORY, horiz=TRUE)
```

Next, let's try to plot a cross-tab of license status vs. license category.

```{r}
ward6detail <- sqldf("select LICENSECATEGORY, LICENSESTATUS, BBL_LICENSE_FACT_ID from bbl where WARD = 6 order by LICENSECATEGORY, LICENSESTATUS")
head(ward6detail)
ward6xtabs <- xtabs(~LICENSESTATUS+LICENSECATEGORY, data=ward6detail)
```

```{r fig.width=10, fig.height=6}
par(las=2, mar=c(2, 12, 2, 2)) # Labels perpendicular to axis, wide left margin
barplot(ward6xtabs, horiz=TRUE, legend=rownames(ward6xtabs))
```

Example of a More Complex Query
-------------------------------
```{r}
query <- 
"select BBL_LICENSE_FACT_ID as bbl_id, 
LATITUDE as lat, LONGITUDE as lon, 
WARD as ward, LICENSESTATUS as status, 
BUSINESSIMPROVEMENTDISTRICT as bid
from bbl
where LICENSESTATUS != 'CANCELLED'
and WARD != 'NA' 
and LATITUDE <> 0 
AND LONGITUDE <> 0"

results <- sqldf(query)
```

```{r fig.width=12, fig.height=24}
library(ggplot2)
qplot(lon, lat, data=results, facets=ward~status, col=bid)
```
